Psychophysics of musical elements in the discrete-time representation of sound

### 2014-12-22 article/article.tex
arrumada autoria para primeiro upload
retirado o caminho para as figuras do artigo para compatibilizar com o esperado pelo arxiv

\author{Renato Fabbri}
 \homepage{http://www.ifsc.usp.br/~fabbri}
 \email{renato.fabbri@gmail.com}
  \affiliation{ 
Instituto de F\'isica de S\~ao Carlos, Universidade de S\~ao Paulo (IFSC/USP)
}

\author{Vilson Vieira da Silva Jr.}
  \homepage{http://automata.cc/}
  \email{vilson@void.cc}
 \altaffiliation[Also at ]{IFSC-USP}

\author{Antônio Pessoti}
  \homepage{}
  \email{debcris.cor@gmail.com}
 \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC-USP}

\author{D\'ebora Corr\^ea}
  \homepage{}
  \email{debcris.cor@gmail.com}
 \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC-USP}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/pcm}
        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is represented by 25 samples with 4 bits each.}
        \label{fig:PCM}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/waveForms_}
    \caption{Basic musical waveforms: (a) the synthetic  waveforms; (b) the real waveforms.}
        \label{fig:formasDeOnda}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/waveSpectrum}
    \caption{Spectrum of basic artificial musical waveforms.}
        \label{fig:espectroDeOndas}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/oboeNaturalSampledSpectrum}
    \caption{Spectrum of the sonic waves of a natural oboe note and from a sampled period. The natural sound has fluctuations in the harmonics and in its noise, while the sampled period note has a perfectly harmonic spectrum.}
        \label{fig:espectroOboe}
\end{figure*}

@@

 \begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{figures/amostras2c__}
     \caption{Oscillation of 2 samples (maximum frequency for any $f_a$). The first coefficient reflects a detachment (\emph{offset} or \emph{bias}) and the second coefficient specifies the oscillation amplitude.}
         \label{fig:amostras2}
 \end{figure*}

@@

 \begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{figures/amostras3b}
     \caption{Three fixed samples presents only one non-null frequency. $c_1=c_2^*$ and $w_1 \equiv w_2$.}
         \label{fig:amostras3}
 \end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/amostras4___}
    \caption{Frequency components for 4 samples.}
        \label{fig:amostras4}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/amostras4formas__}
    \caption{Basic wave forms with 4 samples.}
        \label{fig:formas4}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/amostras6_}
    \caption{Frequency components for 6 samples: 4 sinusoids, one of them is the \emph{bias} with zero frequency.}
        \label{fig:amostras6}
\end{figure*}

@@

Figure~\ref{fig:amostras6} presents the sinusoidal decomposition for 6 samples and figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
In this case, the waveforms have spectra with fundamental differences: square and triangular have the same components but with different proportions, while the sawtooth have an extra component.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/amostras6formas__}
    \caption{Basic waveforms with 6 samples: triangular and square waveforms have odd harmonics, with different proportions and phases; the sawtooth has even harmonics.}
        \label{fig:formas6}
\end{figure*}

@@

\begin{figure}[h!]
    \centering
        \includegraphics[width=.5\textwidth]{figures/espacializacao___}
    \caption{Detection of sound source localization: schema used to calculate Interaural Time Difference (ITD) and Interaural Intensity Difference (IID).}
    \label{fig:spac}
\end{figure}

Considering only the direct incidences in each ear, the equations are quite simple. An object placed at $(x,y)$, as in figure~\ref{fig:spac}, is distant of each ear by:

@@

Figure~\ref{fig:spac} depicts the acoustic shadow of the cranium, an important phenomenon to perception of source azimuthal angle in the cone of confusion. The cone itself is not shown in figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature. Given the filtering and diffraction dependent of the sound spectrum, it is hard, if not impossible, to correctly draw the confusion cone. Even so, the cone of confusion can be understood as a cone with its top placed in the middle of head and growing out in the direction of each ear~\cite{hrtf}.

@@

\begin{figure}[h!]
    {\centering
        \includegraphics[width=\columnwidth]{figures/mixagem}}
    \caption{Mixing of three sound sequences. The amplitudes are directly overlapped.}
        \label{fig:mixagem}
\end{figure}

@@

\begin{figure}[h!]
{    \centering
        \includegraphics[width=\columnwidth]{figures/concatenacao}}
    \caption{Concatenation of three sound sequences by temporal overlap of its samples.}
        \label{fig:concatenacao}
\end{figure}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/lut}
    \caption{Search in the \emph{lookup table} to synthesize sounds in different frequencies using an unique waveform with high resolution.}
        \label{fig:lut}                                                                                                            
\end{figure*}

@@

\begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{figures/transicao}
     \caption{Intensity transitions to different values of $\alpha$ (see equations~\ref{seqAmp} and ~\ref{transAmp}).}
         \label{fig:transicao}
\end{figure}

@@

\begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{figures/convolucao}
     \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde of the other sequence.}
         \label{fig:conv}
\end{figure*}

@@

An important property is the time shift caused by convolution with a shifted impulse. Despite computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound.
In figure~\ref{fig:delays} it is possible to observe the shift caused by convolution with an impulse. Depending on the intensity of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies). In the latter case, the process resembles granular synthesis, reverbs and equalization.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/delays}
    \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Disposed in increasing order of its pulse density.}
        \label{fig:delays}
\end{figure*}

@@

There are many methods and tools to obtain IIR filters. The text bellow lists a selection of them for didactic purposes and as a reference. They are well behaved filters whose aspects are described in figure~\ref{fig:iir}.

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/iir}
    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of equations~\ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} and \ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
        \label{fig:iir}
\end{figure*}

\begin{enumerate}
  \item Low-pass with a simple pole, module of the frequency response in the upper left corner of figure~\ref{fig:iir}. The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which occurs an attenuation of $3dB$.
  The coefficients $a_0$ and $b_1$ of the IIR filter are given by means of the intermediate variable $x \in [e^{-\pi},1]$:

@@

  \item High-pass filter with a simple pole,  module of its frequency responses at the upper right corner of figure~\ref{fig:iir}. The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of the intermediate variable $x \in [e^{-\pi},1]$:

@@

The band-pass filter in the lower left corner of figure~\ref{fig:iir} has the following coefficients:

@@

The coefficients of band-reject filter, depicted in the lower right of figure~\ref{fig:iir}, are:

@@

\begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{figures/ruidos}
     \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: spectrum and waveforms.}
         \label{fig:ruidos}
\end{figure*}

@@

The noise from equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
\ref{eq:azul} and \ref{eq:violeta} are presented in figure~\ref{fig:ruidos}. As stated above, the spectra were built with the same phase and frequency for each coefficient, making it straightforward to observe the contribution of treble harmonics and bass frequencies.

@@

\begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{figures/vibrato}
     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave,  considered $f_s=44.1kHz$.}
         \label{fig:vibrato}
\end{figure}

@@

\begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{figures/tremolo}
     \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (considered a sample frequency of $f_s=44.1kHz$).}
         \label{fig:tremolo}
\end{figure*}

@@

\subsubsection{Convolution for rhythm and meter}
It is possible to use a musical pulse -- such as specified by a BPM
tempo -- and to place an impulse at the start of each beat, with the purpose of
establishing metric and rhythms: the
convolution with an impulse shifts the sound 
to impulse position.
For example, two impulses equally spaced builds a binary division into the
pulse. Two signals, one with 2 impulses and the other with 3 impulses, both 
equally spaced in the pulse duration, yields a pulse
maintenance with a rhythmic in which eases both binary or ternary
divisions of the pulse, and this is found in many ethnic and traditional musical styles~\cite{Gramani}. 
Absolute values of the impulses implyes in proportions among the
amplitudes of the sonic reincidences.
The use of convolution with impulses in this context is explored in the
music piece \emph{Trenzinho de caipiras impulsivos}. These features
embrace the creation of `sound amalgams' based on granular synthesis, for now refer specially to the
figure~\ref{fig:pulsoSubAgl}. This piece is a link to next section contents. 
The source code of the music piece is online, as part of the \massa\ toolbox\cite{MASSA}.

@@

\noindent with $\Lambda_X=\lfloor \Delta . f_s \rfloor\;\;\forall\;\; X \; \in
(A,D,R,\;)$ and being $\xi$ a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ or $\;\xi=10^{\frac{-40}{20}}=10^{-2}$. The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}. The stated second in equations~\ref{eq:adsr} attend both introduction and ending of sound from zero intensity because they are linear. Schematically, figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations. For example, between attack and decay it is possible to add an extra partition where the maximum amplitude remains for more than a peak. Another common example is the use of more elaborated tracings of attack or decay. The music piece \emph{ADa e SaRa}, available 
in \massa, explores many configurations of the ADSL envelope.

@@

\begin{figure}[htpq!]
    \centering
        \includegraphics[width=\columnwidth]{figures/adsr}
    \caption{An ADSR envelope (\emph{Attack, Decay, Sustain, Release}) applied
        to an arbitrary sound sequence. The linear variation of the amplitude is
        above, in blue. Below the amplitude variation is exponential.}
        \label{fig:adsr}
\end{figure}

@@

The elements of $S_j$ can be overlapped by mixing them together, as in
equation~\ref{eq:mixagem} and figure~\ref{fig:mixagem}, for building intervals
and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of the events
in $S_j$, as in equation~\ref{eq:concatenacao} and in figure ~\ref{fig:concatenacao},
builds melodic sequences and rhythms, which are associated with the musical
`horizontal thought'. The fundamental frequency $f$ and the starting moment
(attack) are generally the most important characteristics of the elements $s_j$ in
$S_j$. This makes possible the creation of music made by pitches (both
harmony and melody) and by the presence of temporal metrics and rhythms.

@@

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{figures/movContraponto}
    \caption{Different motions of counterpoint aiming to preserve independence
        between voices. There are 3 types of motion: direct, contrary and
        oblique. The parallel motion is a type of direct motion.}
        \label{fig:movContraponto}
\end{figure}

@@

The pulse is subdivided into equal parts and is also repeated in sequence.
These relations (division and concatenation) usually follow relations of small
order integers. The occurrences, in written and ethnic
music, of musical pulse division (and their sequential grouping at
time),  are, in ascendent disposition: 2, 4 and 8, after that 3, 6 (two groups of 3 or 3 groups of 2) and 9 and
12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, complementing
1-9 and 12. Other metrics are less common, like division and grouping in 13,
17, etc, and are mainly used in contexts of experimental music and classical
music of XX and XXI. No matter how complex they seem, metrics are common
compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
A schematic exposition is shown in figure~\ref{fig:pulsoSubAgl}.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{figures/metricaMusical}
    \caption{Divisions and groupings of the musical pulse for establishing a
        metric. Divisions of the quarter note, regarded as the
        pulse, is presented in the left. The time signature yielded by
        groupings of the music pulse is presented in the right.}
        \label{fig:pulsoSubAgl}
\end{figure*}

@@

Each level $j$ has some sections $i$. When $i$ has three different
values (or multiple of three) there is a perfect relation. When $i$ has only
two, four or eight different values, than there is an imperfect relation,
as shown in figure~\ref{fig:pulsoSubAgl}.

@@

The arcs can be decomposed in two convergent sequences: the first reaches the
apex and the second returns from apex to start region.
This apex is called climax by traditional music theory. It is
possible to differ between arcs whose climax are placed at the beginning, middle, end,
and at the first and second half of the considered duration. These structures are
shown in figure~\ref{fig:climax}. The varying parameter can be unexistent, a case in which
the arc consists only of a reference structure~\cite{Schoenberg}.

\begin{figure}
    \centering
        \includegraphics[width=\columnwidth]{figures/climax}
        \caption{Canonical distinctions of musical climax in a given melody and
        other domains. Considered possibilities are: climax at beginning,
        climax at first half, climax at middle, climax at second half and climax
        at the end. The x and y-axis is not properly specified since the parameters
        can be non-existent, as in a reference structure.}
        \label{fig:climax}
\end{figure}

>>>>

\author{Renato Fabbri}
 \email{renato.fabbri@gmail.com}
  \affiliation{ 
Instituto de F\'isica de S\~ao Carlos, Universidade de S\~ao Paulo (IFSC/USP)
}

\author{Vilson Vieira da Silva Junior}
  \homepage{http://automata.cc/}
  \email{vilson@void.cc}
 \altaffiliation[Also at ]{IFSC-USP}

\author{Antônio Carlos Silvano Pessotti}
  \email{antoniopessotti@gmail.com}
 \altaffiliation[Instituto de Estudos da Linguagem ]{IEL/UNICAMP}

\author{D\'ebora Cristina Corr\^ea}
  \email{debcris.cor@gmail.com}
 \altaffiliation[Instituto de Ci\^encias Matem\'aticas e de Computa\c{c}\~ao ]{ICMC/USP}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{pcm}
        \caption{Pulse Code Modulation (PCM) audio: an analogical signal is represented by 25 samples with 4 bits each.}
        \label{fig:PCM}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{waveForms_}
    \caption{Basic musical waveforms: (a) the synthetic  waveforms; (b) the real waveforms.}
        \label{fig:formasDeOnda}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{waveSpectrum}
    \caption{Spectrum of basic artificial musical waveforms.}
        \label{fig:espectroDeOndas}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{oboeNaturalSampledSpectrum}
    \caption{Spectrum of the sonic waves of a natural oboe note and from a sampled period. The natural sound has fluctuations in the harmonics and in its noise, while the sampled period note has a perfectly harmonic spectrum.}
        \label{fig:espectroOboe}
\end{figure*}

@@

 \begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{amostras2c__}
     \caption{Oscillation of 2 samples (maximum frequency for any $f_a$). The first coefficient reflects a detachment (\emph{offset} or \emph{bias}) and the second coefficient specifies the oscillation amplitude.}
         \label{fig:amostras2}
 \end{figure*}

@@

 \begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{amostras3b}
     \caption{Three fixed samples presents only one non-null frequency. $c_1=c_2^*$ and $w_1 \equiv w_2$.}
         \label{fig:amostras3}
 \end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{amostras4___}
    \caption{Frequency components for 4 samples.}
        \label{fig:amostras4}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{amostras4formas__}
    \caption{Basic wave forms with 4 samples.}
        \label{fig:formas4}
\end{figure*}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{amostras6_}
    \caption{Frequency components for 6 samples: 4 sinusoids, one of them is the \emph{bias} with zero frequency.}
        \label{fig:amostras6}
\end{figure*}

@@

Figure~\ref{fig:amostras6} presents the sinusoidal decomposition for 6 samples and Figure~\ref{fig:formas6} presents the decomposition of the basic wave forms.
In this case, the waveforms have spectra with fundamental differences: square and triangular have the same components but with different proportions, while the sawtooth have an extra component.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{amostras6formas__}
    \caption{Basic waveforms with 6 samples: triangular and square waveforms have odd harmonics, with different proportions and phases; the sawtooth has even harmonics.}
        \label{fig:formas6}
\end{figure*}

@@

\begin{figure}[h!]
    \centering
        \includegraphics[width=.5\textwidth]{espacializacao___}
    \caption{Detection of sound source localization: schema used to calculate Interaural Time Difference (ITD) and Interaural Intensity Difference (IID).}
    \label{fig:spac}
\end{figure}

Considering only the direct incidences in each ear, the equations are quite simple. An object placed at $(x,y)$, as in Figure~\ref{fig:spac}, is distant of each ear by:

@@

Figure~\ref{fig:spac} depicts the acoustic shadow of the cranium, an important phenomenon to perception of source azimuthal angle in the cone of confusion. The cone itself is not shown in Figure~\ref{fig:spac} because it is not exactly a cone and its precise dimensions were not encountered in the literature. Given the filtering and diffraction dependent of the sound spectrum, it is hard, if not impossible, to correctly draw the confusion cone. Even so, the cone of confusion can be understood as a cone with its top placed in the middle of head and growing out in the direction of each ear~\cite{hrtf}.

@@

\begin{figure}[h!]
    {\centering
        \includegraphics[width=\columnwidth]{mixagem}}
    \caption{Mixing of three sound sequences. The amplitudes are directly overlapped.}
        \label{fig:mixagem}
\end{figure}

@@

\begin{figure}[h!]
{    \centering
        \includegraphics[width=\columnwidth]{concatenacao}}
    \caption{Concatenation of three sound sequences by temporal overlap of its samples.}
        \label{fig:concatenacao}
\end{figure}

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{lut}
    \caption{Search in the \emph{lookup table} to synthesize sounds in different frequencies using an unique waveform with high resolution.}
        \label{fig:lut}                                                                                                            
\end{figure*}

@@

\begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{transicao}
     \caption{Intensity transitions to different values of $\alpha$ (see equations~\ref{seqAmp} and ~\ref{transAmp}).}
         \label{fig:transicao}
\end{figure}

@@

\begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{convolucao}
     \caption{Graphical interpretation of convolution. Each resulting sample is the sum of the previous samples of a signal, with each one multiplied by the retrograde of the other sequence.}
         \label{fig:conv}
\end{figure*}

@@

An important property is the time shift caused by convolution with a shifted impulse. Despite computationally expensive, it is possible to create \emph{delay lines} by means of sound convolution with an impulse response that has an impulse for each reincidence of the sound.
In Figure~\ref{fig:delays} it is possible to observe the shift caused by convolution with an impulse. Depending on the intensity of the impulses, the result is perceived as rhythm (from an impulse for each couple of seconds to about 20 impulses per second) or as pitch (from about 20 impulses per second and higher frequencies). In the latter case, the process resembles granular synthesis, reverbs and equalization.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{delays}
    \caption{Convolution with the impulse: shifting (a), delay lines (b) and granular synthesis~(c). Disposed in increasing order of its pulse density.}
        \label{fig:delays}
\end{figure*}

@@

There are many methods and tools to obtain IIR filters. The text bellow lists a selection of them for didactic purposes and as a reference. They are well behaved filters whose aspects are described in Figure~\ref{fig:iir}.

@@

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{iir}
    \caption{Modules for the frequency response (a), (b), (c) and (d) for IIR filters of equations~\ref{eq:passa-baixas}, \ref{eq:passa-altas}, \ref{eq:passa-banda} and \ref{eq:rejeita-banda} respectively, considering different cutoff frequencies, center frequencies and band width.}
        \label{fig:iir}
\end{figure*}

\begin{enumerate}
  \item Low-pass with a simple pole, module of the frequency response in the upper left corner of Figure~\ref{fig:iir}. The general equation has the cutoff frequency $f_c \in (0,\frac{1}{2})$, fraction of the sample frequency $f_s$ in which occurs an attenuation of $3dB$.
  The coefficients $a_0$ and $b_1$ of the IIR filter are given by means of the intermediate variable $x \in [e^{-\pi},1]$:

@@

  \item High-pass filter with a simple pole,  module of its frequency responses at the upper right corner of Figure~\ref{fig:iir}. The general equation with cutoff frequency $f_c \in (0,\frac{1}{2})$ is calculated by means of the intermediate variable $x \in [e^{-\pi},1]$:

@@

The band-pass filter in the lower left corner of Figure~\ref{fig:iir} has the following coefficients:

@@

The coefficients of band-reject filter, depicted in the lower right of Figure~\ref{fig:iir}, are:

@@

\begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{ruidos}
     \caption{Colors of noise generated by equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom}, \ref{eq:azul} and \ref{eq:violeta}: spectrum and waveforms.}
         \label{fig:ruidos}
\end{figure*}

@@

The noise from equations~\ref{eq:branco}, \ref{eq:rosa}, \ref{eq:marrom},
\ref{eq:azul} and \ref{eq:violeta} are presented in Figure~\ref{fig:ruidos}. As stated above, the spectra were built with the same phase and frequency for each coefficient, making it straightforward to observe the contribution of treble harmonics and bass frequencies.

@@

\begin{figure}[h!]
     \centering
         \includegraphics[width=\columnwidth]{vibrato}
     \caption{Spectrogram of a sound with a sinusoidal vibrato of $3Hz$ and one octave of depth in a $1000Hz$ sawtooth wave,  considered $f_s=44.1kHz$.}
         \label{fig:vibrato}
\end{figure}

@@

\begin{figure*}
     \centering
         \includegraphics[width=\textwidth]{tremolo}
     \caption{Tremolo with a depth of $V_{dB}=12dB$, with a sawtooth waveform as its oscillatory pattern, with $f'=1.5Hz$ in a sine of $f=40Hz$ (considered a sample frequency of $f_s=44.1kHz$).}
         \label{fig:tremolo}
\end{figure*}

@@

\subsubsection{Convolution for rhythm and meter}
It is possible to use a musical pulse -- such as specified by a BPM
tempo -- and to place an impulse at the start of each beat, with the purpose of
establishing metric and rhythms: the
convolution with an impulse shifts the sound 
to impulse position.
For example, two impulses equally spaced builds a binary division into the
pulse. Two signals, one with 2 impulses and the other with 3 impulses, both 
equally spaced in the pulse duration, yields a pulse
maintenance with a rhythmic in which eases both binary or ternary
divisions of the pulse, and this is found in many ethnic and traditional musical styles~\cite{Gramani}. 
Absolute values of the impulses implyes in proportions among the
amplitudes of the sonic reincidences.
The use of convolution with impulses in this context is explored in the
music piece \emph{Trenzinho de caipiras impulsivos}. These features
embrace the creation of `sound amalgams' based on granular synthesis, for now refer specially to the
Figure~\ref{fig:pulsoSubAgl}. This piece is a link to next section contents. 
The source code of the music piece is online, as part of the \massa\ toolbox\cite{MASSA}.

@@

\noindent with $\Lambda_X=\lfloor \Delta . f_s \rfloor\;\;\forall\;\; X \; \in
(A,D,R,\;)$ and being $\xi$ a small value that provides a satisfactory \emph{fade in} and \emph{fade out}, e.g.\ $\xi=10^{\frac{-80}{20}}=10^{-4}\;$ or $\;\xi=10^{\frac{-40}{20}}=10^{-2}$. The lower the $\xi$, the slower the \emph{fade}, like the $\alpha$ illustrated in Figure~\ref{fig:transicao}. The stated second in equations~\ref{eq:adsr} attend both introduction and ending of sound from zero intensity because they are linear. Schematically, Figure~\ref{fig:adsr} shows the ADSR envelope in a classical implementation that supports many variations. For example, between attack and decay it is possible to add an extra partition where the maximum amplitude remains for more than a peak. Another common example is the use of more elaborated tracings of attack or decay. The music piece \emph{ADa e SaRa}, available 
in \massa, explores many configurations of the ADSL envelope.

@@

\begin{figure}[htpq!]
    \centering
        \includegraphics[width=\columnwidth]{adsr}
    \caption{An ADSR envelope (\emph{Attack, Decay, Sustain, Release}) applied
        to an arbitrary sound sequence. The linear variation of the amplitude is
        above, in blue. Below the amplitude variation is exponential.}
        \label{fig:adsr}
\end{figure}

@@

The elements of $S_j$ can be overlapped by mixing them together, as in
equation~\ref{eq:mixagem} and Figure~\ref{fig:mixagem}, for building intervals
and chords. This reflects the `vertical thought' in music. On the other hand, the concatenation of the events
in $S_j$, as in equation~\ref{eq:concatenacao} and in Figure~\ref{fig:concatenacao},
builds melodic sequences and rhythms, which are associated with the musical
`horizontal thought'. The fundamental frequency $f$ and the starting moment
(attack) are generally the most important characteristics of the elements $s_j$ in
$S_j$. This makes possible the creation of music made by pitches (both
harmony and melody) and by the presence of temporal metrics and rhythms.

@@

\begin{figure}[h!]
    \centering
        \includegraphics[width=\columnwidth]{movContraponto}
    \caption{Different motions of counterpoint aiming to preserve independence
        between voices. There are 3 types of motion: direct, contrary and
        oblique. The parallel motion is a type of direct motion.}
        \label{fig:movContraponto}
\end{figure}

@@

The pulse is subdivided into equal parts and is also repeated in sequence.
These relations (division and concatenation) usually follow relations of small
order integers. The occurrences, in written and ethnic
music, of musical pulse division (and their sequential grouping at
time),  are, in ascendent disposition: 2, 4 and 8, after that 3, 6 (two groups of 3 or 3 groups of 2) and 9 and
12 (three and 4 groups of 3). At last, the prime numbers 5 and 7, complementing
1-9 and 12. Other metrics are less common, like division and grouping in 13,
17, etc, and are mainly used in contexts of experimental music and classical
music of XX and XXI. No matter how complex they seem, metrics are common
compositions and decompositions of 1-9 equal parts~\cite{Gramani,Roederer}.
A schematic exposition is shown in Figure~\ref{fig:pulsoSubAgl}.

\begin{figure*}
    \centering
        \includegraphics[width=\textwidth]{metricaMusical}
    \caption{Divisions and groupings of the musical pulse for establishing a
        metric. Divisions of the quarter note, regarded as the
        pulse, is presented in the left. The time signature yielded by
        groupings of the music pulse is presented in the right.}
        \label{fig:pulsoSubAgl}
\end{figure*}

@@

Each level $j$ has some sections $i$. When $i$ has three different
values (or multiple of three) there is a perfect relation. When $i$ has only
two, four or eight different values, than there is an imperfect relation,
as shown in Figure~\ref{fig:pulsoSubAgl}.

@@

The arcs can be decomposed in two convergent sequences: the first reaches the
apex and the second returns from apex to start region.
This apex is called climax by traditional music theory. It is
possible to differ between arcs whose climax are placed at the beginning, middle, end,
and at the first and second half of the considered duration. These structures are
shown in Figure~\ref{fig:climax}. The varying parameter can be unexistent, a case in which
the arc consists only of a reference structure~\cite{Schoenberg}.

\begin{figure}
    \centering
        \includegraphics[width=\columnwidth]{climax}
        \caption{Canonical distinctions of musical climax in a given melody and
        other domains. Considered possibilities are: climax at beginning,
        climax at first half, climax at middle, climax at second half and climax
        at the end. The x and y-axis is not properly specified since the parameters
        can be non-existent, as in a reference structure.}
        \label{fig:climax}
\end{figure}
